Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2023-02-23 paper/paper.tex
Revise section 6

Writing and proving properties about code contracts is difficult because it is a formal process, and because it reasons about constraints that define the entire input space, and the entire output space, of a high-assurance component at once.
%
Case studies have shown that designers, as well as formal method practitioners, make mistakes writing these contracts, especially if the computation or temporal reasoning in the contract is complex.

@@

A test contract defines the test scenario for the code contract to implement.
%
It does this by assuming the test input and guarantying the test output akin to a unit test.
%
The code contract is used as the implementation, \eg, it is the test subject.
%
The \agr\ analysis then proves if the code contract, given the test inputs, is strong enough to prove the test contract outputs.
%
Here the test contract assumptions strengthen the code contract assumptions to a single input, and the test contract guarantees weaken, or leave unchanged, the code contract output for that test input, depending on the goal of the test.
%
In other words, according to Liskov substitution, \agr\ proves if the code contract is a safe substitution for the test contract.

@@

\figref{fig:test} is a unit test, with its test contract, for the monitor in \figref{fig:hardened}.
%
The test contract checks if the monitor output coincides with the response that comes one step after the request.
%
It also checks that the alarm is not output.
%
\begin{compactitem}
\item \lineref{line:tst-index}: An index to use to define the test input and expected test output through time.
\item \lineref{line:tst-assume-1} and \lineref{line:tst-assume-2}: The test input. These define when the request and response message events should appear relative to the index.
\item \lineref{line:tst-guarantee-1} and \lineref{line:tst-guarantee-2}: The expected test output.
The test additionally proves that nothing happens after the initial test input and expected output, \ie, no more input is provided and no more output is generated.
\item \linesref{line:tst-imp-start}{line:tst-imp-end}: The implementation with the component under test (\lineref{line:tst-imp-comp}).
\end{compactitem}
%
\figref{fig:hardened-certificate}(a) is the \agr\ output for the test showing that it passes.

Test contracts are effective in detecting vacuity, under-specified behavior, and missed corner cases.
%
They enable designers to iteratively develop code contracts inside OSATE with \agr, and they build assurance that the code contract computes the desired behavior with black-box, white-box, and other test coverage metrics.
%
These same tests can be synthesized to the backend implementation for traditional testing, if for example, the code contract is a model of an existing legacy binary.

@@

Synthesis maps from model and specifications to code.
%
The \splt\ tool traverses the system architecture looking for occurrences of high-assurance components specified by code contracts;
%
for each such occurrence it generates a \ckml\ program.
%
\splt\ supports three kinds of behavioral specification from which to generate code.
%
In increasing order of expressiveness, these are: regular expressions, contiguity types, and code contracts.
%
In earlier work we have described how regular expressions and contiguity types are used as specification languages for verified filter implementations.
%
In this paper, we extend this \emph{specification oriented} approach to the Turing-complete specifications of code contracts in \agr.
%
We give a short review of that previous work before going into detail on the synthesis.

@@

\paragraph{Contiguity types\cite{contiguity-types}:\/} contiguity types, similar to regular expressions, are effective at describing data at the network message level, \eg, as flat strings.
%
The notation is however much more expressive, being able to declare many kinds of so-called \emph{self-describing} messages, \ie, those where message structure is dependent on data held elsewhere in the message.
%
\figref{fig:filter-spec} gives a contiguity type specification for the well-formed waypoints of our example system.
%
We have formalized and verified a contiguity type matcher, which has been extensively applied throughout our case studies.

@@

When full computational power is needed, we can use the expressive power of \agr\ code contracts to write arbitrarily complex filter or monitor operations, as we have already seen.
%
Our formal model of \agr\ syntax and semantics is applied to extract code contracts and map them to equivalent logic functions.
%
In particular, given a code contract with functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics given in \secref{sec:code-contracts} can be applied to define a stream-oriented \emph{step} function $\konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env}$ defined as
%
\[
 \konst{strmStep}\;t \; E = \sem{\mathit{Eqs} \cdot G}^E_t
\]
%
Now suppose ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially imperative.
%
Then we can direcctly create a logic function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
%
corresponding to \konst{strmStep} that takes input and state as parameters and returns an updated state and output.
%
Thus the environment $\mathit{Env}$ holding all previous values of each variable can be replaced by \emph{parameters}, each of which is able to hold only one data element.
%
Assuming $\mathit{E}' = \konst{strmStep}\;t\; \mathit{E}$, we can formally prove that
%
\[
\vdash (\mathit{E}' \vert_{\mathbb{S}},
 \mathit{E}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{E}\vert_{\mathbb{I}(t)},
                   \mathit{E}\vert_{\mathbb{S}(t)})
\]
%
In other words, projecting out the state and output components from the result of $\konst{strmStep}\;t\; \mathit{E}$ equals the result of applying \konst{stateStep} on the input and state components of $\mathit{E}$ at time $t$.
%
This theorem connects the stream semantics of the code contract to a pure logic function over the values appearing in the stream.
%
From this point we can use tools in the \ckml\ eco-system to jump from the setting of higher order logic to \ckml\ abstract syntax \cite{cakeml-translator}.
%
In this way we have moved from stream semantics to logic function to CakeML evaluation semantics in a property-preserving way.
%
Once in \ckml, further necessary features can be added, namely imperative variables for storing state between component invocations, also invocations of the (verified) foreign function interface of \ckml\ \cite{cakeml-monadic}.
%
Finally, the \ckml compiler can be applied to the program.
%
It delivers an executable which can be shown to preserve the original contract.

@@

\subsection{Generating \ckml}
For any of the three specification languages, we take the following general steps to arrive at an executable.
%
\begin{compactenum}
\item Map from specification to computable function in logic, producing a proof of equivalence.
\item Add parsers and printers (specified by contiguity types) for input and output.
\item Map to \ckml\ and add calls to input/output library routines via the \ckml\ foreign function interface.
\item Invoke the \ckml\ compiler.
\end{compactenum}
%
In the following, we examine further details in both filter and monitor synthesis.

@@

A filter is intended to be simple, although it may make deep semantic checks.
%
A filter has one input port and one output; messages on the input that the filter policy admits pass unchanged to the output port;
%
all others are dropped (not passed on).
%
We have investigated two kinds of filter.
%
In the first, a relatively shallow scan of the input suffices to enforce the policy.
%
For example, we have used the expressive power of regular expressions and Contiguity Types \cite{contiguity-types} to enforce \emph{lightweight} bounds constraints on GPS coordinates in UxAS messages.
%
On the other hand, a filter may need to parse the input buffer into a data structure specified in \agr\ and apply a user-defined \emph{wellformedness} property, also specified in \agr, to the data.
%
Arbitrarily complex wellformedness checks can be made in this way.
%
\figref{fig:filter-spec} shows a combination where the checking specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+} depends on an underlying check specified by the contiguity type checking bounds on waypoints.

The verdict of a filter is made and performed within one thread invocation.
%
Thus, in its given time slice, the following steps must be completed:
\begin{compactenum}
\item
The filter checks to see if there is any input available.
%
If there is none then it yields control; otherwise:

\item The input is read (and parsed if need be);

\item The wellformedness predicate is evaluated on the input;

\item If the predicate returns \konst{true} then the input buffer
 is copied to the output, otherwise no action is taken; and

\item The filter yields control.
\end{compactenum}

@@

The contiguity type specification and wellformedness predicate for the filter are shown in \figref{fig:filter-spec} and the synthesized \ckml\ code is in \figref{fig:filter-cakeml}.
%
The code is called at dispatch by the scheduler.
%
The \texttt{API.callFFI} is the link to the communication fabric to capture input and provide output.
%
The body of the function restates the filter contract to make the appropriate assignments in a way that matches the truth value of the predicate in the filter guarantee.

@@

As discussed above, a monitor specification is mapped to a state transformation function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}.
\]
%
which computes as follows:
%
\begin{compactenum}

\item Each input buffer is parsed into data of the type specified by the port type;

\item
New values for the state variables are computed, in dependency order.
%
At this point variables defined by \emph{followed-by} are dealt with.
%
Suppose the variable definitions have the following form:
%
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
%
In the generated function, for the first invocation of \konst{stateStep} only, the initializations are executed in order:
%
\[
  v_1 = i_1; \cdots ;  v_n = i_n;
\]
%
In all later invocations, the \emph{non-initialization} assignments are performed, also in order:
%
\[
  v_1 = e_1; \cdots ; v_n = e_n; \\
\]

\item Values of the outputs are computed;

\item Outputs are written and the new state is written;

\item The monitor yields control.
\end{compactenum}

The \konst{stepFn} for the monitor of the example described in \secref{sec:example} is displayed in \figref{fig:monitor-cakeml}.

@@

\subsection{Incorporation into system execution}

@@

\paragraph{Partiality:\/} partiality is an important consideration for both filters and monitors: the input data might not parse or the computation of a filter could be badly written and fail at runtime.
%
In such cases, the filter should recover and yield control without passing the input onwards.
%
For these failure cases, the filter is behaving as it should, but we must also guard against situations in which a \emph{correctly specified} filter or monitor fails in the runtime.
%
This kind of defect arises when a filter, for example, \emph{ought} to accept a message, but lack of resources results in the filter failing to do so.
%
For example, the parse of a message might need more space than has been allocated;
%
another example could be if the time slice provided by the scheduler is too short for the computation to finish.
%
Thus resource bounds need to be included in the correctness argument.
%
Some preliminary work that could support formal proofs of such bounds has been done for \ckml\ \cite{cakeml-space-cost}.

>>>>

Writing and proving properties about code contracts is difficult because it is a formal process, and because it reasons about constraints that define the entire input space, and the entire output space, of a high-assurance component at once.
%
Our case studies have shown that designers, as well as formal method practitioners, make mistakes writing these contracts, especially if the computation or temporal reasoning in the contract is complex.

@@

A test contract defines the test scenario for the code contract to implement.
%
It does this by assuming the test input and guarantying the test output akin to a unit test.
%
The code contract is used as the implementation, \eg, it is the test subject.
%
The \agr\ analysis then proves if the code contract, given the test inputs, is strong enough to prove the test contract outputs.
%
Here the test contract assumptions strengthen the code contract assumptions to a single input, and the test contract guarantees weaken, or leave unchanged, the code contract output guarantees for that test input, depending on the goal of the test.
%
In other words, according to Liskov substitution, \agr\ proves if the code contract is a safe substitution for the test contract.

@@

\figref{fig:test} is a unit test, with its test contract, for the monitor in \figref{fig:hardened}.
%
The test contract checks if the monitor output coincides with the response that comes one step after the request.
%
It also checks that the alarm is not output.
%
\begin{compactitem}
\item \lineref{line:tst-index}: an index to use to define the test input and expected test output through time.
\item \lineref{line:tst-assume-1} and \lineref{line:tst-assume-2}: the test input. These define when the request and response message events should appear relative to the index.
\item \lineref{line:tst-guarantee-1} and \lineref{line:tst-guarantee-2}: the expected test output.
The test additionally proves that nothing happens after the initial test input and expected output, \ie, no more input is provided and no more output is generated.
\item \linesref{line:tst-imp-start}{line:tst-imp-end}: the implementation with the component under test (\lineref{line:tst-imp-comp}).
\end{compactitem}
%
\figref{fig:hardened-certificate}(a) is the \agr\ output for the test showing that it passes.

Test contracts are effective in detecting vacuity, under-specified behavior, and missed corner cases.
%
They enable designers to iteratively develop code contracts with \agr\ in the modeling environment, and they build assurance that the code contract computes the desired behavior with black-box, white-box, and other test coverage metrics.
%
These same tests can be synthesized to the backend implementation for traditional testing, if for example, the code contract is a model of an existing legacy binary.

@@

Synthesis maps from model and specifications to code.
%
The \splt\ tool traverses the system architecture looking for occurrences of high-assurance components specified by code contracts;
%
for each such occurrence it generates a \ckml\ program.
%
\splt\ supports three kinds of behavioral specification from which to generate code.
%
In increasing order of expressiveness, these are: regular expressions, contiguity types, and code contracts.
%
In earlier work we have described how regular expressions and contiguity types are used as specification languages for verified filter implementations.
%
In this paper, we extend this \emph{specification oriented} approach to the Turing-complete specifications of code contracts in \agr.
%
We give a short review of our previous work before going into detail on the synthesis.

@@

\paragraph{Contiguity types:\/} contiguity types, \cite{contiguity-types}, similar to regular expressions, are effective at describing data at the network message level, \eg, as flat strings.
%
The notation is however much more expressive than regular expressions, being able to declare many kinds of so-called \emph{self-describing} messages, \ie, those where message structure is dependent on data held elsewhere in the message.
%
\figref{fig:filter-spec} gives a contiguity type specification for the well-formed waypoints of our example system.
%
We have formalized and verified a contiguity type matcher, and that matcher has been extensively applied throughout our case studies.

@@

When full computational power is needed, we can use the expressive power of \agr\ code contracts to write arbitrarily complex filter or monitor operations, as we have already seen.
%
Our formal model of \agr\ syntax and semantics is applied to extract code contracts and map them to equivalent logic functions.
%
In particular, given a code contract with functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics given in \secref{sec:code-contracts} can be applied to define a stream-oriented \emph{step} function $\konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env}$ defined as
%
\[
 \konst{strmStep}\;t \; E = \sem{\mathit{Eqs} \cdot G}^E_t
\]
%
Now suppose ${\Eqs}\cdot G$ is sequentially imperative.
%
Then we can direcctly create a logic function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
%
corresponding to \konst{strmStep} that takes input and state as parameters and returns an updated state and output.
%
Thus the environment $\mathit{Env}$ holding all previous values of each variable can be replaced by \emph{parameters}, each of which is able to hold only one data element.
%
Assuming $\mathit{E}' = \konst{strmStep}\;t\; \mathit{E}$, we can formally prove that
%
\[
\vdash (\mathit{E}' \vert_{\mathbb{S}},
 \mathit{E}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{E}\vert_{\mathbb{I}(t)},
                   \mathit{E}\vert_{\mathbb{S}(t)})
\]
%
In other words, projecting out the state and output components from the result of $\konst{strmStep}\;t\; \mathit{E}$ equals the result of applying \konst{stateStep} on the input and state components of $\mathit{E}$ at time $t$.
%
This theorem connects the stream semantics of the code contract to a pure logic function over the values appearing in the stream.
%
From this point we can use tools in the \ckml\ eco-system to jump from the setting of higher order logic to \ckml\ abstract syntax \cite{cakeml-translator}.
%
In this way we have moved from stream semantics to logic function to CakeML evaluation semantics in a property-preserving way.
%
Once in \ckml, further necessary features can be added, namely imperative variables for storing state between component invocations, also invocations of the (verified) foreign function interface of \ckml\ \cite{cakeml-monadic}.
%
Finally, the \ckml\ compiler can be applied to the program.
%
It delivers an executable which can be shown to preserve the original contract.

@@

\subsection{Generating \ckml}
For any of the three specification languages, we take the following general steps to arrive at an executable.
%
\begin{compactenum}
\item Map from specification to computable function in logic, producing a proof of equivalence.
\item Add parsers and printers (specified by contiguity types) for input/output.
\item Map to \ckml\ and add calls to input/output library routines via the \ckml\ foreign function interface.
\item Invoke the \ckml\ compiler.
\end{compactenum}
%
In the following, we examine further details in both filter and monitor synthesis.

@@

A filter is intended to be simple, although it may make deep semantic checks.
%
A filter has one input and one output; messages on the input that the filter policy admits pass unchanged to the output;
%
all others are dropped (not passed on).
%
We have investigated two kinds of filter.
%
In the first, a relatively shallow scan of the input suffices to enforce the policy.
%
For example, we have used regular expressions and Contiguity Types to enforce bounds on GPS coordinates in UxAS messages.
%
On the other hand, a filter may need to parse the input buffer into a data structure specified in \agr\ and apply a user-defined property, also specified in \agr, to the data.
%
Arbitrarily complex checks can be made in this way.
%
\figref{fig:filter-spec} shows a combination where the checking specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+} depends on an underlying check specified by the contiguity type checking bounds on waypoints.

The verdict of a filter is made and performed within one thread invocation.
%
Thus, in its given time slice, the following steps must be completed:
\begin{compactenum}
\item
the filter checks to see if there is any input available.
%
If there is none then it yields control; otherwise:

\item the input is read (and parsed if need be);

\item the wellformedness predicate is evaluated on the input;

\item if the predicate returns \konst{true} then the input buffer
 is copied to the output, otherwise no action is taken; and

\item the filter yields control.
\end{compactenum}

@@

The synthesized \ckml\ code for \figref{fig:filter-spec} is in \figref{fig:filter-cakeml}.
%
The code is called at dispatch by the scheduler.
%
The \texttt{API.callFFI} is the link to the communication fabric to capture input and provide output.
%
The body of the function restates the filter contract to make the appropriate assignments in a way that matches the truth value of the predicate in the filter guarantee.

@@

As discussed previously, the code contract specification for the monitor is mapped to a state transformation function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}.
\]
%
which computes as follows:
%
\begin{compactenum}

\item each input buffer is parsed into data of the type specified by the port type;

\item
new values for the state variables are computed, in dependency order.
%
At this point variables defined by \emph{followed-by} are dealt with.
%
Suppose the variable definitions have the following form:
%
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
%
In the generated function, for the first invocation of \konst{stateStep} only, the initializations are executed in order:
%
\[
  v_1 = i_1; \cdots ;  v_n = i_n;
\]
%
In all later invocations, the \emph{non-initialization} assignments are performed, also in order:
%
\[
  v_1 = e_1; \cdots ; v_n = e_n; \\
\]

\item values of the outputs are computed;

\item outputs are written and the new state is written;

\item the monitor yields control.
\end{compactenum}
\noindent The \konst{stepFn} for the monitor in the example described in \secref{sec:example} is displayed in \figref{fig:monitor-cakeml}.

@@

\subsection{Incorporation into System Execution}

@@

\paragraph{Partiality of the Correctess Argument:\/} partiality is an important consideration for both filters and monitors as the input data might not parse or the computation of a filter could be badly written and fail at runtime.
%
In such cases, the filter should recover and yield control without passing the input onwards.
%
For these failure cases, the filter is behaving as it should, but we must also guard against situations in which a correctly specified filter or monitor fails in the runtime.
%
This kind of defect arises when a filter, for example, ought to accept a message, but lack of resources results in the filter failing to do so.
%
For example, the parse of a message might need more space than has been allocated;
%
another example could be if the time slice provided by the scheduler is too short for the computation to finish.
%
Thus resource bounds need to be included in the correctness argument.
%
Some preliminary work that could support formal proofs of such bounds has been done for \ckml\ \cite{cakeml-space-cost}.
